---
title: "Panel C — Shannon's entropy of the CENP-A/CENH3 alignment"
author: "DToL Centromere Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 14,
  fig.height = 6
)
```

## Overview

This report reproduces the **positional Shannon entropy plot** shown in Figure 1, Panel C.

**Summary:**

- CENP-A / CENH3 sequences from 327 DToL species were aligned with MAFFT and curated manually.
- Perkinsus was removed (outlier divergence).
- Shannon entropy was computed per alignment position using **entropia**, split by broad taxonomic group (Fungi, Invertebrates, Vertebrates, Viridiplantae).
- The analysis lives in `PhylogeneticProfiling/20_cenpa_analysis/20.1-march/`.

**Sequence cross-check vs Panel D tree:** The entropy alignment (`cenpa_leaflabels_manually_march.aligned.fasta`, **428 sequences**) predates the final curated Panel D tree (`cenpa430_H3_archaea10.fasta`, **430 CenpA sequences**). The two extra CenpA sequences were added later during tree curation. All 428 entropy sequences are present in the Panel D tree.

```{r libraries}
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(patchwork)
  library(Biostrings)   # for reading FASTA
  library(ape)          # for read.tree, keep.tip
})
```

```{r paths}
BASE        <- "/home/jg2070/Desktop/dtol_review_August"
CENPA_DIR   <- file.path(BASE, "PhylogeneticProfiling/20_cenpa_analysis/20.1-march")
ENTROPY_DIR <- file.path(CENPA_DIR, "entropia-split")
PUB2_DIR    <- file.path(BASE, "DToL_phylogenomics_publication_325genomes")

# Entropy TSVs are group-level (not affected by removing 2 species)
ENTROPY_TSV  <- file.path(ENTROPY_DIR, "cenpa_entropy_split_phylo.tsv")
UNTRIMMED_TSV<- file.path(ENTROPY_DIR, "cenpa_entropy_split_phylo_coarse_untrimmed.tsv")
ALN_FASTA    <- file.path(CENPA_DIR,  "cenpa_leaflabels_manually_march.aligned.fasta")

# 325-species CENP-A+H3+archaea FASTA and tree
BNNI_FASTA   <- file.path(PUB2_DIR, "04_cenpa_phylogeny",
                           "cenpa430_H3_archaea10.aligned.clipkit.325sp.fasta")
BNNI_TREE    <- file.path(PUB2_DIR, "04_cenpa_phylogeny",
                           "iqtree2_cenpa430_H3_archaea10_325sp_bnni.treefile")
```

---

## 1. Sequence cross-check: entropy alignment vs Panel D tree

```{r sequence_crosscheck}
# Read both FASTA headers
aln_ids <- sub("^>", "", grep("^>", readLines(ALN_FASTA), value = TRUE))
aln_ids <- trimws(sub(" .*", "", aln_ids))

tree_ids_all <- sub("^>", "", grep("^>", readLines(BNNI_FASTA), value = TRUE))
tree_ids_all <- trimws(sub(" .*", "", tree_ids_all))

# Archaea IDs (excluded from CenpA count)
archaea_ids <- c("BAD86478.1", "KKK41979.1", "KXH71038.1", "OIO41945.1",
                 "OIO61677.1", "OLS16336.1", "OLS18261.1", "OLS21974.1",
                 "OLS22332.1", "OLS24873.1")

# H3 IDs (from tree tip labels in H3_all.aligned.fasta)
h3_fasta <- file.path(TREE_DIR, "H3_all.aligned.fasta")
h3_ids   <- sub("^>", "", grep("^>", readLines(h3_fasta), value = TRUE))
h3_ids   <- trimws(sub(" .*", "", h3_ids))

cenpa_tree_ids <- setdiff(tree_ids_all, c(archaea_ids, h3_ids))

cat("Entropy alignment sequences:       ", length(aln_ids), "\n")
cat("CenpA sequences in Panel D tree:   ", length(cenpa_tree_ids), "\n")
cat("Entropy seqs absent from tree:     ",
    length(setdiff(aln_ids, cenpa_tree_ids)), "\n")
cat("Tree CenpA seqs absent from entropy:",
    length(setdiff(cenpa_tree_ids, aln_ids)), "\n")

if (length(setdiff(aln_ids, cenpa_tree_ids)) > 0) {
  cat("\n--- Entropy seqs NOT in Panel D tree ---\n")
  print(setdiff(aln_ids, cenpa_tree_ids))
}
if (length(setdiff(cenpa_tree_ids, aln_ids)) > 0) {
  cat("\n--- Panel D CenpA seqs NOT in entropy alignment ---\n")
  print(head(setdiff(cenpa_tree_ids, aln_ids), 20))
}
```

---

## 2. Load pre-computed entropy data

Shannon entropy values per alignment position per group were computed by **entropia** (tool documented in `entropia_calculate_shannon_entropy.log`).

```{r load_entropy}
ent <- read.delim(ENTROPY_TSV, stringsAsFactors = FALSE, check.names = FALSE)
cat("Entropy table dimensions:", nrow(ent), "positions ×", ncol(ent), "columns\n")
cat("Groups:", paste(setdiff(colnames(ent), "pos"), collapse = ", "), "\n")
```

```{r tidy_entropy}
groups_pal <- c(
  Fungi         = "#6A1B9A",
  Invertebrates = "#E65100",
  Vertebrates   = "#1565C0",
  Viridiplantae = "#2E7D32"
)

ent_long <- ent %>%
  pivot_longer(-pos, names_to = "group", values_to = "entropy") %>%
  mutate(group = factor(group, levels = names(groups_pal)))

head(ent_long)
```

---

## 3. Entropy plot — all groups (Figure 1, Panel C)

```{r entropy_plot_main, fig.width=14, fig.height=5}
p_entropy <- ggplot(ent_long, aes(x = pos, y = entropy, colour = group)) +
  geom_line(linewidth = 0.5, alpha = 0.85) +
  scale_colour_manual(values = groups_pal, name = "Taxonomic group") +
  scale_x_continuous(name = "Alignment position") +
  scale_y_continuous(name = "Shannon entropy (bits)", limits = c(0, 1),
                     breaks = seq(0, 1, 0.25)) +
  labs(
    title    = "Positional Shannon entropy — CENP-A/CENH3 alignment",
    subtitle = sprintf("428 CenpA sequences from 327 DToL species (Perkinsus removed); %d alignment positions",
                       max(ent$pos))
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position  = "right",
    panel.grid.minor = element_blank()
  )

p_entropy
```

---

## 4. Smoothed entropy (rolling average)

```{r entropy_smoothed, fig.width=14, fig.height=5}
roll_mean <- function(x, k = 10) {
  stats::filter(x, rep(1 / k, k), sides = 2)
}

ent_smooth <- ent_long %>%
  group_by(group) %>%
  arrange(pos) %>%
  mutate(entropy_smooth = as.numeric(roll_mean(entropy, k = 10))) %>%
  ungroup()

p_smooth <- ggplot(ent_smooth, aes(x = pos, y = entropy_smooth, colour = group)) +
  geom_line(linewidth = 0.7, alpha = 0.90, na.rm = TRUE) +
  scale_colour_manual(values = groups_pal, name = "Taxonomic group") +
  scale_x_continuous(name = "Alignment position") +
  scale_y_continuous(name = "Shannon entropy (10-pos rolling avg)", limits = c(0, 1)) +
  labs(title    = "Smoothed positional entropy (window = 10 positions)") +
  theme_bw(base_size = 12) +
  theme(legend.position = "right", panel.grid.minor = element_blank())

p_smooth
```

---

## 5. Per-group entropy summary statistics

```{r entropy_summary}
ent_long %>%
  group_by(group) %>%
  summarise(
    mean_entropy   = mean(entropy,  na.rm = TRUE),
    median_entropy = median(entropy, na.rm = TRUE),
    sd_entropy     = sd(entropy,    na.rm = TRUE),
    pct_high       = mean(entropy > 0.5, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  knitr::kable(caption = "Per-group positional entropy summary")
```

---

## 6. Coarse structural regions (untrimmed alignment)

The TSV from the coarse structural-region split provides entropy per broad structural domain.

```{r entropy_regions, fig.width=14, fig.height=5}
if (file.exists(UNTRIMMED_TSV)) {
  ent_u <- read.delim(UNTRIMMED_TSV, stringsAsFactors = FALSE, check.names = FALSE)
  ent_u_long <- ent_u %>%
    pivot_longer(-pos, names_to = "group", values_to = "entropy") %>%
    mutate(group = factor(group, levels = names(groups_pal)))

  ggplot(ent_u_long, aes(x = pos, y = entropy, colour = group)) +
    geom_line(linewidth = 0.5, alpha = 0.85) +
    scale_colour_manual(values = groups_pal, name = "Taxonomic group") +
    scale_x_continuous(name = "Alignment position (untrimmed)") +
    scale_y_continuous(name = "Shannon entropy (bits)", limits = c(0, 1)) +
    labs(title = "Untrimmed alignment — coarse structural regions") +
    theme_bw(base_size = 12) +
    theme(legend.position = "right", panel.grid.minor = element_blank())
} else {
  cat("Untrimmed TSV not found:", UNTRIMMED_TSV, "\n")
}
```

---

## 7. Save publication figure

```{r save_fig, eval=FALSE}
OUT <- file.path(dirname(knitr::current_input(dir = TRUE)), "output")
dir.create(OUT, showWarnings = FALSE)

ggsave(file.path(OUT, "panel_C_entropy.pdf"), p_entropy, width = 14, height = 5)
ggsave(file.path(OUT, "panel_C_entropy.png"), p_entropy, width = 14, height = 5, dpi = 300)
ggsave(file.path(OUT, "panel_C_entropy_smoothed.pdf"), p_smooth, width = 14, height = 5)
cat("Saved to:", OUT, "\n")
```

---

## Session info

```{r session}
sessionInfo()
```
