---
title: "Panel D — CenpA + H3 + Archaea phylogeny (bnni converged tree)"
author: "DToL Centromere Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 16,
  fig.height = 16
)
```

## Overview

This report reproduces the **CenpA + H3 + Archaea phylogeny** shown in Figure 1, Panel D.

**Key features:**
- **Tree:** IQ-TREE2 (-bnni converged) — `iqtree2_cenpa430_H3_archaea10_bnni.treefile` (1337 tips: 430 CenpA, 897 H3, 10 archaea)
- **Model:** Q.pfam+R7 (best-fit)
- **UFBoot convergence:** Achieved in ≤1000 replicates with `-bnni` (zero non-convergence warnings vs prior run); run time ~38h
- **Rooting:** Archaea MRCA (outgroup)
- **Centromere categories displayed:**
  - Satellite = #ff006e
  - Transposon = #3a86ff
  - Holocentric = #2d7d32 (dark green), displayed as **triangles**
  - Satellite/transposon = #8338ec
  - **Cryptic** = Unknown + Monocentric sequence unknown (combined = #808080)
  - H3 = not shown (no symbol)
  - Archaea = not shown (no symbol)
- **iTOL annotation files** are in `19_curated_tree/iTOL_bnni/`

```{r libraries}
suppressPackageStartupMessages({
  library(ape)
  library(ggtree)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
})
```

```{r paths}
BASE      <- "/home/jg2070/Desktop/dtol_review_August"
PROF_DIR  <- file.path(BASE, "PhylogeneticProfiling")
PUB2_DIR  <- file.path(BASE, "DToL_phylogenomics_publication_325genomes")

# 325-species versions (daLycEuro1.1 + mRhiSin1.1 removed)
TREE_DIR      <- file.path(PUB2_DIR, "04_cenpa_phylogeny")
ITOL_DIR      <- file.path(TREE_DIR, "iTOL_bnni")
BNNI_TREEFILE <- file.path(TREE_DIR, "iqtree2_cenpa430_H3_archaea10_325sp_bnni.treefile")
CLASSIF_TSV   <- file.path(PUB2_DIR, "02_ASR", "centromere_classification_clean.tsv")
CODES_CSV     <- file.path(PUB2_DIR, "04_cenpa_phylogeny", "expected_325_codes.txt")
H3_FASTA      <- file.path(PROF_DIR, "19_curated_tree", "H3_all.aligned.fasta")
```

---

## 1. IQ-TREE2 convergence summary

The original UFBoot run (without `-bnni`) reached a correlation coefficient of **0.970** and showed an oscillating pattern suggesting non-convergence. The `-bnni` run optimised each bootstrap replicate with NNI moves and converged on the **first 1000 replicates** (zero non-convergence warnings), completing in ~38 hours.

| Run | UFBoot correlation | Non-convergence warnings | Runtime |
|-----|-------------------|--------------------------|---------|
| Without `-bnni` | 0.970 (oscillating) | Many | ~120h |
| With `-bnni` | ≥0.99 (flat) | 0 | ~38h |

**IQ-TREE2 command:**
```
/home/jg2070/miniforge3/envs/forORTHO/bin/iqtree2 \
  -s cenpa430_H3_archaea10.aligned.clipkit.fasta \
  -B 1000 -bnni -nm 2000 \
  -T 12 --prefix iqtree2_cenpa430_H3_archaea10_bnni
```

---

## 2. Load tree and metadata

```{r load_tree}
tr_raw <- read.tree(BNNI_TREEFILE)
cat("Tree tips:", Ntip(tr_raw), "\n")
cat("Is rooted:", is.rooted(tr_raw), "\n")

archaea_ids <- c("BAD86478.1", "KKK41979.1", "KXH71038.1", "OIO41945.1",
                 "OIO61677.1", "OLS16336.1", "OLS18261.1", "OLS21974.1",
                 "OLS22332.1", "OLS24873.1")

# Root at archaea MRCA
tr <- root(tr_raw,
           node         = getMRCA(tr_raw, archaea_ids),
           resolve.root = TRUE)
cat("Rooted at archaea MRCA. Tips:", Ntip(tr), "\n")
```

```{r load_metadata}
classif <- read.table(CLASSIF_TSV, sep = "\t", header = TRUE,
                      stringsAsFactors = FALSE)
classif$classification <- trimws(classif$classification)

codes <- read.csv(CODES_CSV, header = FALSE,
                  col.names = c("status", "fasta", "group", "genus", "species"),
                  stringsAsFactors = FALSE)
codes$prefix <- sub("\\..*", "", codes$fasta)
codes$group  <- trimws(codes$group)

h3_ids <- trimws(sub(" .*", "", sub("^>", "",
  grep("^>", readLines(H3_FASTA), value = TRUE))))

cat("H3 IDs loaded:", length(h3_ids), "\n")
cat("CenpA species classified:", nrow(classif), "\n")
```

---

## 3. Classify tips

```{r classify_tips}
tips <- tr$tip.label

# Classification: merge Unknown + Monocentric → "Cryptic"
classify_tip <- function(tid) {
  if (tid %in% archaea_ids) return("Archaea")
  if (tid %in% h3_ids)      return("H3")
  pfx <- sub("_.*", "", tid)
  cls <- classif$classification[classif$species_code == pfx]
  if (!length(cls))
    cls <- classif$classification[sub("[0-9]+$", "", classif$species_code) == pfx]
  if (length(cls) == 1) return(cls)
  "Unknown"
}

tip_cat_raw <- vapply(tips, classify_tip, character(1))

# Recode Unknown + Monocentric sequence unknown → Cryptic
tip_cat <- dplyr::recode(tip_cat_raw,
  "Unknown"                      = "Cryptic",
  "Monocentric sequence unknown" = "Cryptic"
)

cat("Tip classification summary:\n")
print(sort(table(tip_cat), decreasing = TRUE))
```

---

## 4. Colour palette and shape map

```{r palette}
# Colour palette matching the existing DToL iTOL annotation
pal_cenpa <- c(
  "Satellite"          = "#ff006e",
  "Transposon"         = "#3a86ff",
  "Holocentric"        = "#2d7d32",   # dark green
  "Satellite/transposon" = "#8338ec",
  "Cryptic"            = "#808080",   # Unknown + Monocentric
  "H3"                 = "#BDBDBD",
  "Archaea"            = "#212121"
)

# Shapes: Holocentric = triangle (17), all others = circle (16)
shape_map <- c(
  "Satellite"          = 16,
  "Transposon"         = 16,
  "Holocentric"        = 17,   # triangle
  "Satellite/transposon" = 16,
  "Cryptic"            = 16,
  "H3"                 = 16,
  "Archaea"            = 16
)

tip_df <- data.frame(
  label    = tips,
  category = tip_cat,
  colour   = pal_cenpa[tip_cat],
  shape    = shape_map[tip_cat],
  stringsAsFactors = FALSE
)
```

---

## 5. Taxonomic group per tip (for clade highlights)

```{r taxonomy}
get_group <- function(pfx) {
  g <- codes$group[codes$prefix == pfx]
  if (!length(g)) g <- codes$group[sub("[0-9]+$", "", codes$prefix) == pfx]
  if (length(g)) g[1] else ""
}

tip_pfx <- sub("_.*", "", tips)
tip_grp <- vapply(tip_pfx, get_group, character(1))

ins_grps  <- c("Coleoptera", "Hymenoptera", "Diptera", "Lepidoptera", "Trichoptera",
               "Plecoptera", "Ephemeroptera", "Odonata", "Psocodea", "Hemiptera",
               "Blattodea", "Dermaptera", "Neuroptera")
vert_grps <- c("Aves", "Mammalia", "Actinopterygii", "Reptilia", "Tunicata")
fung_grps <- "Fungi"
plnt_grps <- c("Dicots", "Monocots", "Bryophyta", "Algae")

is_h3_tip   <- tips %in% h3_ids
is_arch_tip <- tips %in% archaea_ids

cenpa_ins  <- tips[tip_grp %in% ins_grps   & !is_h3_tip & !is_arch_tip]
cenpa_vert <- tips[tip_grp %in% vert_grps  & !is_h3_tip & !is_arch_tip]
cenpa_fung <- tips[tip_grp %in% fung_grps  & !is_h3_tip & !is_arch_tip]
cenpa_plnt <- tips[tip_grp %in% plnt_grps  & !is_h3_tip & !is_arch_tip]

cat("CenpA: insects =", length(cenpa_ins), " vertebrates =", length(cenpa_vert),
    " fungi =", length(cenpa_fung), " plants =", length(cenpa_plnt), "\n")
```

---

## 6. Identify clade MRCA nodes (purity-based)

We use pre-computed high-purity clade nodes (found by iterating over all internal nodes and selecting the one with the highest proportion of CenpA from the target group, subject to recall ≥ 50%).

```{r clade_nodes}
safe_mrca <- function(tset) {
  tset <- tset[tset %in% tr$tip.label]
  if (length(tset) < 2) return(NA_integer_)
  getMRCA(tr, tset)
}

# Find high-purity clade nodes (re-derives from tree topology)
find_purity_node <- function(target_tips, all_cenpa_tips, tr,
                              min_purity = 0.6, min_recall = 0.50) {
  target_tips <- intersect(target_tips, tr$tip.label)
  all_cenpa_tips <- intersect(all_cenpa_tips, tr$tip.label)
  n_target <- length(target_tips)
  if (n_target < 2) return(NA_integer_)
  n_nodes <- tr$Nnode
  n_tips  <- Ntip(tr)
  best_f1 <- -1; best_node <- NA_integer_
  for (nd in (n_tips + 1):(n_tips + n_nodes)) {
    desc <- tryCatch(phangorn::Descendants(tr, nd, type = "tips")[[1]], error = function(e) integer(0))
    if (!length(desc)) next
    desc_labels <- tr$tip.label[desc]
    n_desc      <- length(desc_labels)
    tp    <- sum(desc_labels %in% target_tips)
    fp    <- sum(desc_labels %in% all_cenpa_tips) - tp
    purity <- tp / (tp + fp + 1e-10)
    recall <- tp / n_target
    if (purity < min_purity || recall < min_recall) next
    f1 <- 2 * purity * recall / (purity + recall + 1e-10)
    if (f1 > best_f1) { best_f1 <- f1; best_node <- nd }
  }
  best_node
}

all_cenpa_tips <- tips[!is_h3_tip & !is_arch_tip]

cat("Searching for purity-based clade nodes (may take a moment)...\n")
suppressPackageStartupMessages(library(phangorn))

node_ins  <- find_purity_node(cenpa_ins,  all_cenpa_tips, tr)
node_vert <- find_purity_node(cenpa_vert, all_cenpa_tips, tr)
node_fung <- find_purity_node(cenpa_fung, all_cenpa_tips, tr)
node_plnt <- find_purity_node(cenpa_plnt, all_cenpa_tips, tr)

cat(sprintf("Clade nodes — insects: %d  vertebrates: %d  fungi: %d  plants: %d\n",
            node_ins, node_vert, node_fung, node_plnt))
```

---

## 7. Build circular ggtree plot

```{r ggtree_plot, fig.width=17, fig.height=17}
p <- ggtree(tr, layout = "circular", linewidth = 0.10, colour = "grey60") %<+% tip_df +
  # Background H3 / archaea tips: small, faint
  geom_tippoint(
    aes(colour = category,
        size   = ifelse(category %in% c("H3", "Archaea"), 0.25, 0.75),
        alpha  = ifelse(category %in% c("H3", "Archaea"), 0.40, 1.00),
        shape  = shape),
    show.legend = FALSE
  ) +
  # CenpA tips on top: larger, opaque, correct shape
  geom_tippoint(
    data = function(d) subset(d, !is.na(category) &
                                !grepl("^H3$|^Archaea$", category)),
    aes(colour = category, shape = shape),
    size  = 1.0,
    alpha = 0.95
  ) +
  scale_colour_manual(
    values = pal_cenpa,
    name   = "Centromere architecture",
    labels = c(
      "Satellite"           = "Satellite",
      "Transposon"          = "Transposon",
      "Holocentric"         = "Holocentric",
      "Satellite/transposon"= "Satellite/Transposon",
      "Cryptic"             = "Cryptic (Unknown/Monocentric)",
      "H3"                  = "H3",
      "Archaea"             = "Archaea"
    ),
    na.value = "#BDBDBD"
  ) +
  scale_shape_identity() +
  scale_size_identity() +
  scale_alpha_identity() +
  guides(colour = guide_legend(
    override.aes = list(size = 3, shape = c(16, 16, 17, 16, 16, 16, 16))
  )) +
  theme(
    legend.position   = c(0.87, 0.45),
    legend.key.size   = unit(0.35, "cm"),
    legend.text       = element_text(size = 7.5),
    legend.title      = element_text(size = 8.5, face = "bold"),
    legend.background = element_rect(fill = "white", colour = "grey80", linewidth = 0.3),
    plot.title        = element_text(size = 10, face = "bold", hjust = 0.5)
  ) +
  ggtitle("CenpA + H3 + Archaea phylogeny  (bnni converged, 1337 tips)")

# Add clade highlights and arc labels
clade_info <- list(
  list(node = node_ins,  label = "CenpA Insects",      col = "#E65100"),
  list(node = node_vert, label = "CenpA Vertebrates",   col = "#1565C0"),
  list(node = node_fung, label = "CenpA Fungi",         col = "#6A1B9A"),
  list(node = node_plnt, label = "CenpA Viridiplantae", col = "#2E7D32")
)

for (ci in clade_info) {
  if (!is.na(ci$node)) {
    p <- p +
      geom_hilight(node = ci$node, fill = ci$col, alpha = 0.08, extend = 0.4) +
      geom_cladelab(node = ci$node, label = ci$label,
                    fontsize = 2.5, offset = 0.5, offset.text = 0.08,
                    barsize = 0.6, barcolour = ci$col, textcolour = ci$col,
                    hjust = 0.5)
  }
}

p
```

---

## 8. iTOL annotation files

Pre-generated iTOL annotation files are in `19_curated_tree/iTOL_bnni/`:

| File | Type | Description |
|------|------|-------------|
| `cenpa_H3_archaea_bnni_rooted.nwk` | Tree | Rooted Newick — upload to iTOL |
| `iTOL_centromere_symbols.txt` | DATASET_SYMBOL | Circles for all CenpA; colours by architecture |
| `iTOL_centromere_architecture.txt` | DATASET_COLORSTRIP | Colour strip by architecture |
| `iTOL_taxonomic_clade.txt` | DATASET_COLORSTRIP | Colour strip by organism group |
| `iTOL_centromere_binary.txt` | DATASET_BINARY | One column per architecture type |
| `iTOL_taxonomic_treecolors.txt` | TREE_COLORS | Tip label colours by taxonomic group |

```{r check_itol}
itol_files <- list.files(ITOL_DIR, full.names = FALSE)
cat("iTOL annotation files available:\n")
for (f in sort(itol_files)) cat(" ", f, "\n")
```

### Symbol colour map

```{r itol_symbol_map}
sym_df <- data.frame(
  Category = c("Satellite", "Transposon", "Holocentric",
               "Satellite/transposon", "Cryptic (Unknown + Monocentric)",
               "H3", "Archaea"),
  Colour   = c("#ff006e", "#3a86ff", "#2d7d32",
               "#8338ec", "#808080",
               "—", "—"),
  Symbol   = c("Circle", "Circle", "Circle",
               "Circle", "Circle",
               "None", "None"),
  stringsAsFactors = FALSE
)
knitr::kable(sym_df, caption = "iTOL DATASET_SYMBOL colour and shape map")
```

---

## 9. Save plots

```{r save_plots, eval=FALSE}
OUT <- file.path(dirname(knitr::current_input(dir = TRUE)), "output")
dir.create(OUT, showWarnings = FALSE)

ggsave(file.path(OUT, "panel_D_cenpa_phylogeny.pdf"), p, width = 17, height = 17)
ggsave(file.path(OUT, "panel_D_cenpa_phylogeny.png"), p, width = 17, height = 17, dpi = 180)
cat("Saved to:", OUT, "\n")
```

---

## Session info

```{r session}
sessionInfo()
```
